### 2025-12-07 manuscript_ten-simple-rules-writing-retreat.tex
fixed typos and minor adjustments before ms submission

% Current address notes
\textcurrency Current Address: Marine Mammal Institute, Department of Fisheries, Wildlife and Conservation Sciences, Oregon State University, Newport, OR, USA

@@

As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023. In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, as each student leads their own research. In addition, given our diverse disciplines, field sites, and lab locations, there was a shared social disconnect within our programme. A writing retreat was planned to not only facilitate community building within our PhD cohort ~\cite{omeara2017, bernery2022}, but to also structure a productive week as we worked toward our dissertation goals ~\cite{mcgrail2006, kornhaber2016}. The retreat was a success both in terms of writing produced as well as connections built between peers sharing a similar PhD journey. The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.

Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats. Although the authors were all PhD students at the time they participated in the retreats, these rules can be applied broadly to any research-oriented or academically-minded group. We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se. Many helpful resources on academic writing exist (e.g. ~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}.

@@

Use the proposal to seek support. Academic researchers may access departmental or divisional funding to support these opportunities for development. There may be external funding available to provide supplemental support if internal money is not available to cover the entire costs of the retreat. This funding could be sought from industry and community organisations (e.g. Lions and Rotary Clubs), or through fundraisers. A well-developed proposal lays a solid foundation for organising the retreat and securing the support needed for success.

@@

Create a schedule for your writing retreat, as it provides the structure for focused writing and thinking time ~\cite{murray2009, tremblay2021}. Use available resources to guide your planning (e.g. ~\cite{tremblay2021}), including this guide (see also Appendices). At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.

@@

We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group. Goals may be quantitative (e.g. word count, page number) or qualitative (e.g. specific sections or revisions). For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts. By identifying a specific writing target ahead of time, friction related to making decisions is reduced during the retreat with goals clear in mind. Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.

The pre-retreat meeting should be held within two weeks of departure to support timely preparation while preserving participant momentum. This timeframe supports goal-oriented preparation while preserving participant momentum. If possible, a casual, in-person meeting is preferred to provide a comfortable setting for participants to meet and connect. Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g., meal coordination, daily facilitator), or ideas for social activities.

@@

Schedule academic discussions around participants' work in a structured manner. A writing retreat gathers your academic peers in a collaborative environment with minimal external distractions. It presents a unique opportunity to garner advice and perspectives on written work (see also Rule 8) or any work-in-progress that participants may be developing. We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat. Access to a projector and screen in our second and third retreats was beneficial. However, previously, we have simply shared from our laptop screens or consulted notes and found it equally engaging due to the small group attending the retreats. Discussions were informal, and we aimed not to exceed 20 minutes per session. It is important to emphasise the casualness of such an exercise---students should not be using their time for writing anxiously preparing a conference-level presentation. These are simply opportunities to share, learn, and exchange insights.

While we found academic discussions materalised naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees. The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.' Undertaking any large-writing project (e.g., a PhD thesis) is a daunting and varying experience for everyone. It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}. A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers. Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students ~\cite{omeara2017, bernery2022}.

@@

Providing feedback is just as valuable as receiving it. This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed. When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components for example, structure, clarity, and flow. If you are an expert in the content, detailed feedback can also be extremely valuable.

>>>>

% Current address notes
\textcurrency Current Address: Geospatial Ecology of Marine Megafauna Laboratory, Marine Mammal Institute, Oregon State University, Newport, OR, USA \& Department of Fisheries, Wildlife, and Conservation Sciences, Oregon State University, Corvallis, OR, USA

@@

As a cohort of PhD students in the Department of Marine Science at the University of Otago (Aotearoa/New Zealand), we organised a 5-day writing retreat at a remote field station in 2023. In New Zealand, PhDs follow research-only programs; as such, we do not participate in coursework that might promote group cohesion, as each student leads their own research. In addition, our diverse disciplines, field sites, and lab locations present challenges in maintaining social cohesion within our programme. A writing retreat was planned to not only facilitate community building within our PhD cohort ~\cite{omeara2017, bernery2022}, but to also structure a productive week as we worked toward our dissertation goals ~\cite{mcgrail2006, kornhaber2016}. The retreat was a success both in terms of writing produced as well as connections built between peers sharing a similar PhD journey. The retreat's success motivated us to organise a second retreat the following year; the second retreat's success inspired a third.

Based on our experience, we present Ten Simple Rules for organising effective (student-led) writing retreats. Although the authors were all PhD students at the time they participated in the retreats, these rules can be applied broadly to any research-oriented or academically-minded group. We outline steps to support the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and post- (Rule 10) writing retreat actions, but we do not cover writing techniques per se. Many helpful resources on academic writing exist (e.g., ~\cite{turbek2016, hotaling2020, carter2020, cargill2021}), including articles in this series ~\cite{weinberger2015, mensh2017, peterson2018}.

@@

Use the proposal to seek support. Academic researchers may access departmental or divisional funding to support these opportunities for development. There may be external funding available to provide supplemental support if internal money is not available to cover the entire costs of the retreat. This funding could be sought from industry and community organisations (e.g., Lions and Rotary Clubs), or through fundraisers. A well-developed proposal lays a solid foundation for organising the retreat and securing the support needed for success.

@@

Create a schedule for your writing retreat, as it provides the structure for focused writing and thinking time ~\cite{murray2009, tremblay2021}. Use available resources to guide your planning (e.g., ~\cite{tremblay2021}), including this guide (see also Appendices). At our university, we collaborated with the Higher Education Development Centre, which provided guidance on designing productive days and structuring the week for our first and second retreats.

@@

We suggest that participants arrive at the pre-retreat meeting with a preliminary goal to share with the group. Goals may be quantitative (e.g., word count, page number) or qualitative (e.g., specific sections or revisions). For reference, during our 5-day retreats, participants averaged 1.2 pages or 550 words per day, working on everything from detailed revisions to drafting full manuscripts. By identifying a specific writing target ahead of time, friction related to making decisions is reduced during the retreat with goals clear in mind. Sharing goals also helps participants learn about each other's research and assess whether goals are realistic.

The pre-retreat meeting should be held within two weeks of departure to support timely preparation while preserving participant momentum. If possible, a casual, in-person meeting is preferred to provide a comfortable setting for participants to meet and connect. Additional objectives may include selecting writing prompts, presentation topics, icebreakers, roles (e.g., meal coordination, daily facilitator), or ideas for social activities.

@@

Schedule academic discussions around participants' work in a structured manner. A writing retreat gathers your academic peers in a collaborative environment with minimal external distractions. It presents a unique opportunity to garner advice and perspectives on written work (see also Rule 8) or any work-in-progress that participants may be developing. We suggest scheduling these discussions after writing blocks are complete for the day and spreading them out over the retreat. Access to a projector and screen in our second and third retreats was beneficial. However, previously, we have simply shared from our laptop screens or consulted notes and found it equally engaging due to the small group attending the retreats. Discussions were informal, and we aimed not to exceed 20 minutes per session. It is important to emphasise the casualness of such an exercise---students should not anxiously prepare a conference-level presentation at the expense of their writing time. These are simply opportunities to share, learn, and exchange insights.

While we found academic discussions materialised naturally throughout the retreats, especially within groups of similarly focused researchers, we also allocated a small block of time each day (see Rule 3) specifically for students to share a talk or lead a discussion with the other attendees. The casual, non-judgemental environment allowed presenting students a chance to practice a seminar talk, gain insight on methodologies, analyses and results, or simply share `tips and tools of the trade.' Undertaking any large-writing project (e.g., a PhD thesis) is a daunting and varying experience for everyone. It is easy to become so engrossed in your work that you forget you are surrounded by people in a similar situation ~\cite{wilson2019}. A writing retreat provides the opportunity to share your own research or discuss scientific/academic topics amongst your peers. Importantly, these sessions offer a chance to get to know one another and connect as fellow PhD students ~\cite{omeara2017, bernery2022}.

@@

Providing feedback is just as valuable as receiving it. This practice encourages a collaborative approach to writing that builds confidence and resilience in both the reviewer and the reviewed. When giving feedback, use neutral language ~\cite{parsons2021} and focus on writing components, such as structure, clarity, and flow. If you are an expert in the content, detailed feedback can also be extremely valuable.
